---
title: "KS Test"
author: Paul Jeffries
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: TRUE
    toc_depth: 2
---

**NOTE: this is an early work in progress. Check back shortly for new additions**

Sources:

https://www.kaggle.com/kemical/kickstarter-projects/home

https://flowingdata.com/2012/05/15/how-to-visualize-and-compare-distributions/

http://www.sthda.com/english/wiki/ggplot2-density-plot-quick-start-guide-r-software-and-data-visualization


```{r setup, echo = F}
# setting the appropriate working directory
knitr::opts_knit$set(root.dir = '~/Desktop/Personal/personal_code/distributional_tests/')
```

```{r results='hide', warning=FALSE, message=FALSE}
# first a few general set-up items / housekeeping items
# setting scipen options to kill all use of scientific notation
options(scipen = 999)

# basic packages needed throughout
library(plyr) # always load prior to dplyr if needed
library(dplyr) # for piping
library(ggplot2) # for visualization
library(ggthemes) # for custom visualization

```

```{r}
# need to get the other newer portion of this dataset
test_df <- read.csv("data/ks-projects-201612.csv")
```

```{r}
head(test_df)
```

The distributional tests we can run here span a wide variety of options. Right now, I'm thinking the below:

- How do the distributions of $ amount funded (dollar-denominated) differ by offering country?
- Does this phenomenon vary based on the category or sub-category of the offering?

Things to think about:

- should I round to nearest high 10, as done below?
- any other filters I'm leaving out?

```{r}
test_df_forviz <- test_df %>% 
  # filters to just kickstarters from the US and GB
  dplyr::filter(country %in% c('US','GB')) %>%
  # filters to just kickstarters under a certain category
  dplyr::filter(category == 'Music') %>%
  # filter to only the cases where the USD-denominated pledged amount was less than 10000
  # this dampens the long tails problem in a non-fancy way that I'll change later
  # dplyr::filter(as.numeric(as.character(usd.pledged)) < 10000) %>%
  mutate(
    # rounded pledged dollar values to nearest 10
    rounded_usdpledged = plyr::round_any(as.numeric(as.character(usd.pledged)), 10)) 
```

```{r}
# ggplot code for the basid density plot
ggplot(data = test_df_forviz, aes(x=as.numeric(as.character(usd.pledged)))) +
  geom_density(aes(fill=factor(country)),alpha = 0.4) +
  scale_x_continuous(limits = c(0,5000)) +
  theme(legend.position = "top") +
  labs(
    title = paste0("PDF for music category kickstarters"),
    y = "Concentration Density",
    x = "Amount Pledged (converted to USD)",
    fill = "Country of origin for kickstarter"
  )

```




```{r}

```




